Prediction of *PIK3CA* mutations from cancer gene expression data

### 2020-07-16 plosOne/plosOne.tex
proof reading

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Prediction of PIK3CA mutations from cancer gene expression data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jun Kang\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}},
Ahwon Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}},
Youn Soo Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}}\textsuperscript{*}\\
\bigskip
\textbf{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}}Department of Hospital Pathology, Seoul St.~Mary's Hospital, College of
Medicine, The Catholic University of Korea, Seoul, South Korea\\
\bigskip
* Corresponding author: lys9908@catholic.ac.kr\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Breast cancers with PIK3CA mutations can be treated with PIK3CA
inhibitors in hormone receptor-positive HER2 negative subtypes. We
applied a supervised elastic net penalized logistic regression model to
predict PIK3CA mutations from gene expression data. This regression
approach was applied to predict modeling using the TCGA pan-cancer
dataset. Approximately 10,845 cases were available for PIK3CA mutation
and mRNA expression data. In 10-fold cross-validation, the model with
\(\lambda\) = 0.01 and \(\alpha\) = 1.0 (Ridge regression) showed the
best performance, in terms of area under the receiver operating
characteristic (AUROC). The final model was developed with selected
hyper-parameters using the entire training set. The training set AUROC
was 0.93, and the test set AUROC was 0.84. The area under the
precision-recall (AUPR) of the training set was 0.66, and the test set
AUPR was 0.39. Cancer types were the most important predictors. Both
nsulin Like Growth Factor 1 Receptor (IGF1R) and the phosphatase and
tensin homolog (PTEN) were the most significant genes in gene expression
predictors. Our study suggests that predicting genomic alterations using
gene expression data is possible, with good outcomes.

@@

Machine learning approaches can be applied to detect genomic alterations
. Machine learning algorithms can build prediction models from a large
number of predictors, such as radiomic features {[}3{]}, pathology image
{[}4{]} or gene expression data {[}5{]}. Because most direct genomic
tests are more specific and sensitive than predictive models, machine
learning approaches may have limited roles in clinical practice,
however, machine learning approaches are ideal when direct tests are
unavailable or fail.

RAS pathway activation predictions have been performed using gene
expression data {[}7{]}. Authors used data from The Cancer Genome Atlas
(TCGA), with a supervised elastic net penalized logistic regression
classifier, with stochastic gradient descent. Their model performance
was 84\% with an area under the receiver operating characteristic
(AUROC) curve, and 63\% with an area under the precision-recall (AUPR)
curve. Importantly, these authors suggested their approach could be
applied to other genomic alterations.

Breast cancer expressing PIK3CA mutations can be treated using PIK3CA
inhibitors, in hormone receptor-positive HER2 negative subtypes {[}8{]}.
The PIK3CA mutation is the second most common driver mutation after
TP53, and is most frequently detected in endometrial carcinoma (45\%),
followed by breast invasive carcinoma (24\%), cervical squamous cell
carcinoma, endo-cervical adenocarcinoma (20\%) and colon adenocarcinoma
(16\%) {[}9{]}.

PIK3CA encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase that
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to generate phosphatidylinositol
3,4,5-triphosphate (PIP\textsubscript{3}). The phosphatase and tensin
homolog (PTEN) converts PIP\textsubscript{2} to PIP\textsubscript{3} in
contrast to PI3K. PIP\textsubscript{3} is a second messenger that
activates protein kinase B (AKT), which is a serine/threonine-specific
protein kinase. AKT inhibits apoptosis and promotes cell proliferation
{[}10{]}.

We applied a supervised elastic net penalized logistic regression model
to predict PIK3CA mutations. We wanted to ascertain whether this
prediction model approach could be applied not only to RAS pathway
activation, but also to PIK3CA mutation predictions across several
cancer types.

@@

We used the TCGA pan-cancer dataset. TCGA archives the following; exome
sequencing, gene expression, DNA methylation, protein expression, and
clinical data from \textgreater{} 10,000 cancer samples across 33 common
cancer types. The TCGA dataset is publically available. PIK3CA mutation
data was extracted using cgdsr rpackage {[}11{]}. Gene expression data
was downloaded from the National Cancer Institute (NCI)'s Genomic Data
Commons (GDC) website. This archives data for TCGA
(https://gdc.cancer.gov/about-data/publications/pancanatlas). Gene
expression in the TCGA pan-cancer dataset is batch-corrected with
normalization.

The target variable was PIK3CA mutation status. PIK3CA status was
considered positive when the case had the following PIK3CA variants
(C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R,
H1047Y) which were the target variants of the Therascreen PIK3CA RGQ PCR
Kit, Qiagen, Hilden, Germany. This kit was approved as a companion
diagnostics test to treat with PIK3CA inhibitor by the United States
Food and Drug Administration.

@@

To narrow down potential predictors, genes with a large median absolute
deviation (\textgreater{} third-quartiles) were selected. Thirty three
cancer type dummy variables were included in predictor variables. We
split three-quarters of the dataset into the training set and one
quarter into the test set. Yeo-Johnson transformation was performed to
correct skewness. Centering and scaling were also performed. All
preprocessing was performed using the recipe r package {[}12{]}.
Penalized logistic regression was applied to prediction modeling.
Ten-fold cross-validation with target variable stratification was
performed over the hyper-parameter grid: \(\lambda\) \{\(10^{-5}\),
\(10^{-4}\),\(10^{-3}\),\(10^{-2}\),\(10^{-1}\), \(10^{0}\)\},
\(\alpha\) \{0.0, 0.25, 0.5, 0.75\}. Lambda (\(\lambda\)) is a penalty
scaling parameter and alpha (\(\alpha\)) is a mixing parameter of
penalty function
(\((1-\alpha)/2 \lVert\beta\rVert_2^2+\alpha\lVert \beta \rVert_1\))
{[}13{]}.

\hypertarget{accessing-model-performance}{%
\subsection{Accessing model
performance}\label{accessing-model-performance}}

Model performance was evaluated using AUROC and AUPR curve approaches.
The AUPR approach is more informative than AUROC for imbalanced datasets
{[}14{]}. The modeling process and accessing model performance were
performed with the tidymodels rpackage {[}15{]}.

@@

10,845 cases were available for both PIK3CA mutation and mRNA expression
data. 5,128 out of 20,502 genes were included in the modeling process,
after filtering for median absolute deviation, as described in the
modeling process method. The prevalence rate for PIK3CA was 0.11 in all
cases. The PIK3CA prevalence rate in each cancer type varied. The median
prevalence rate of PIK3CA for each cancer type was 0.03 (range 0--0.33)
(Figure 1).

@@

Because PIK3CA mutation prevalence varied across cancer types, the
performance of each cancer type was investigated. The AUROC and AUPR
were positively correlated between the training sets and test sets in
cancer type sub-analysis (Figure 2B). The AUPR was high in cancer types
with high PIK3CA mutation rates such as colon, breast and uterus cancer
types. The AUROC did not correlate with PIK3CA mutation rates of each
cancer type (Figure 2C).

@@

The top 30 important predictors are shown (Figure 3). The coefficient is
the parameter of the predictor which represents the effect of the
predictor on prediction. Insulin Like Growth Factor 1 Receptor (IGF1R)
mRNA expression was the strongest negative predictor, and PTEN was the
strongest positive predictor. Both IGF1R and PTEN are key players in the
tyrosine kinase pathway . The cancer types were important predictors.
Some cancer types including uterine carcinosarcoma (UCS), bladder
urothelial carcinoma (BLCA), pancreatic adenocarcinoma (PAAD), lymphoid
neoplasm diffuse large B-cell lymphoma (DLBC) were the strongest
predictors.

@@

Our model showed good performance in predicting PIK3CA mutations in
various cancer types. Our data suggested that the supervised elastic net
penalized logistic regression model could be applied not only to the RAS
activation pathway, but also to other genomic alterations. Both the RAS
activation pathway and PIK3CA mutations are key, common cancer genomic
alterations. When mutated, they exert significant effect on gene
expression in cells. However, the supervised elastic net penalized
logistic regression model cannot be generalized or applied to other
genomic alterations. It might be a challenging prediction of genomic
alterations that are infrequent or have a weak effect on gene
expression.

Prediction modeling from the TCGA pan-cancer dataset can be limiting in
terms of data preprocessing. The gene expression data is processed by
between-sample normalization to remove batch effects. If the model has
been trained from between-sample normalization, a new sample cannot be
exactly preprocessed with normalization which was done on trainset. A
model based on gene expression from the TCGA pan-cancer dataset is
limited in terms of data preprocessing. It is necessary to develop a
preprocessing method that is independent of a dataset, to apply gene
expression data to the prediction model.

Our PIK3CA prediction model was similar to the RAS activation prediction
model in terms of AUROC (0.84). However the AUPR of our model was lower
than the RAS activation model (0.39 versus 0.63). The reason for our
lower AUPR may be explained by the low prevalence rate of PIK3CA
mutations, and an imbalanced dataset. The model for RAS activation
trained with cancer types with more than 0.05 prevalence of RAS
activation to avoid imbalance classification problem. We included all
cancer types in our modeling process. The lower prevalence rate of
target variables meant our dataset had a lower AUPR baseline. In the
sub-analysis performance of each cancer type, the cancer types with
higher PIK3CA mutation rates showed better AUPRs.

Our model included cancer types as predictors, and they were stronger
predictors than gene expression. The varying prevalence of PIC3CA
mutations across cancer types may be a reason for the strong predictive
power of cancer types. If the cancer type was wrong or could not be
determined, our model performance was poor.

Some significant gene expression predictors were closely related to PTEN
and the PI3K pathway. PTEN and IGFR1R were the strongest gene expression
predictors, which has negative and positive predictive powers. IGF1R is
a tyrosine kinase receptor that activates PI3K {[}16{]}. {]}, and PTEN
is an important regulator of PIP\textsubscript{3} by dephosphorylating
PIP\textsubscript{3}, in contrast to PI3K {[}10{]}.

Several studies have attempted to predict genomic alterations from gene
expression data {[}6,17{]}. A study investigated PIK3CA mutation
predictions using gene-expression signatures which is a sum of the
average of the logarithmic gene expression . The model showed good
performance AUROC 0.71 in an independent test set {[}6,17{]}.Another
study predicted copy number alterations with gene expression, using a
multinomial logistic regression model with least absolute shrinkage and
selection operator (LASSO) parameters {[}18{]}. The prediction of the
1p/19q codel was very good, with an AUROC of 0.997, and gene-level
predictions were good, with an AUROC of 0.75 {[}18{]}. A Hidden Markov
Model Approach for Prediction of Genomic Alterations from Gene
Expression Profiling {[}19{]}. A logistic regression model was used for
MYCN gene amplification in neuroblastoma {[}20{]}.

Our study suggested that the prediction of genomic alterations using
gene expression data was possible, with good performance. However,
improved performances are required for clinical tests, and the
standardization of generation processing of gene expression data is also
needed .

@@

\begin{itemize}
\item
  Figure 1. Prevalence rate of PIK3CA mutations across cancer types.
  Cancer type abbreviations are explained in the S1 Appendix.
\item
  Figure 2. Summary of modeling results. (A) Left: receiver operating
  characteristic (ROC) curve. Right: precision-recall (PR) curve of
  training set and test set. The horizontal green line is the PIK3CA
  mutation rate (0.11) (B) Correlation between training set and test set
  of the area under the receiver operating characteristic curve (AUROC),
  and the area under the precision-recall curve (AUPR) among cancer
  types. The gray band is the 95\% confidence interval. Abbreviations
  are explained in the S1 Appendix. (C) Correlations between the PIK3CA
  mutation rate of the AUROC, and the AUPR.
\item
  Figure 3. Coefficient model. (A) Top 30 high mRNA coefficients. (B)
  Cancer type coefficients. Cancer types abbreviations are explained in
  the S1 Appendix.
\end{itemize}

@@

\leavevmode\hypertarget{ref-way_2018_machine_cellreports}{}%
7. Way GP, Sanchez-Vega F, La K, Armenia J, Chatila WK, Luna A, et al.
Machine Learning Detects Pan-cancer Ras Pathway Activation in The Cancer
Genome Atlas. Cell Reports. Elsevier; 2018;23: 172--180.e3.
doi:\href{https://doi.org/10.1016/j.celrep.2018.03.046}{10.1016/j.celrep.2018.03.046}

\leavevmode\hypertarget{ref-andre_2019_alpelisib_nengljmeda}{}%
8. André F, Ciruelos E, Rubovszky G, Campone M, Loibl S, Rugo HS, et al.
Alpelisib for PIK3CA-Mutated, Hormone ReceptorPositive Advanced Breast
Cancer. New England Journal of Medicine. Massachusetts Medical Society;
2019;380: 1929--1940.
doi:\href{https://doi.org/10.1056/NEJMoa1813904}{10.1056/NEJMoa1813904}

\leavevmode\hypertarget{ref-baileyComprehensiveCharacterizationCancer2018}{}%
9. Bailey MH, Tokheim C, Porta-Pardo E, Sengupta S, Bertrand.
Comprehensive Characterization of Cancer Driver Genes and Mutations.
Cell. 2018;173: 371--385.e18.
doi:\href{https://doi.org/10.1016/j.cell.2018.02.060}{10.1016/j.cell.2018.02.060}

\leavevmode\hypertarget{ref-cantleyPhosphoinositide3KinasePathway2002}{}%
10. Cantley LC. The Phosphoinositide 3-Kinase Pathway. Science. American
Association for the Advancement of Science; 2002;296: 1655--1657.
doi:\href{https://doi.org/10.1126/science.296.5573.1655}{10.1126/science.296.5573.1655}

\leavevmode\hypertarget{ref-jacobsenCgdsrRbasedAPI2019}{}%
11. Jacobsen A, Luna A. Cgdsr: R-based API for accessing the MSKCC
cancer genomics data server (CGDS). 2019.

\leavevmode\hypertarget{ref-kuhnRecipesPreprocessingTools2020}{}%
12. Kuhn M, Wickham H. Recipes: Preprocessing tools to create design
matrices. 2020.

\leavevmode\hypertarget{ref-friedmanRegularizationPathsGeneralized2010}{}%
13. Friedman J, Hastie T, Tibshirani R. Regularization Paths for
Generalized Linear Models via Coordinate Descent. J Stat Softw. 2010;33:
1--22.

\leavevmode\hypertarget{ref-saitoPrecisionRecallPlotMore2015}{}%
14. Saito T, Rehmsmeier M. The Precision-Recall Plot Is More Informative
than the ROC Plot When Evaluating Binary Classifiers on Imbalanced
Datasets. PLOS ONE. Public Library of Science; 10: e0118432.
doi:\href{https://doi.org/10.1371/journal.pone.0118432}{10.1371/journal.pone.0118432}

\leavevmode\hypertarget{ref-kuhnTidymodelsEasilyInstall2020}{}%
15. Kuhn M, Wickham H. Tidymodels: Easily install and load the
'tidymodels' packages. 2020.

\leavevmode\hypertarget{ref-leroithInsulinlikeGrowthFactor2003}{}%
16. LeRoith D, Roberts CT. The insulin-like growth factor system and
cancer. Cancer Letters. 2003;195: 127--137.
doi:\href{https://doi.org/10.1016/S0304-3835(03)00159-9}{10.1016/S0304-3835(03)00159-9}

\leavevmode\hypertarget{ref-cizkovaGeneExpressionProfiling2010}{}%
17. Cizkova M, Cizeron-Clairac G, Vacher S, Susini A, Andrieu C,
Lidereau R, et al. Gene Expression Profiling Reveals New Aspects of
PIK3CA Mutation in ERalpha-Positive Breast Cancer: Major Implication of
the Wnt Signaling Pathway. PLOS ONE. Public Library of Science; 5:
e15647.
doi:\href{https://doi.org/10.1371/journal.pone.0015647}{10.1371/journal.pone.0015647}

\leavevmode\hypertarget{ref-muCNAPEMachineLearning2019}{}%
18. Mu Q, Wang J. CNAPE: A Machine Learning Method for Copy Number
Alteration Prediction from Gene Expression. IEEE/ACM Transactions on
Computational Biology and Bioinformatics. 2019; 1--1.
doi:\href{https://doi.org/10.1109/TCBB.2019.2944827}{10.1109/TCBB.2019.2944827}

\leavevmode\hypertarget{ref-gengHiddenMarkovModel2008}{}%
19. Geng H, Ali HH, Chan WC. A Hidden Markov Model Approach for
Prediction of Genomic Alterations from Gene Expression Profiling. In:
Măndoiu I, Sunderraman R, Zelikovsky A, editors. Bioinformatics Research
and Applications. Berlin, Heidelberg: Springer; 2008. pp. 414--425.
doi:\href{https://doi.org/10.1007/978-3-540-79450-9_38}{10.1007/978-3-540-79450-9\_38}

\leavevmode\hypertarget{ref-heGeneSignaturesAssociated2020}{}%
20. He X, Qin C, Zhao Y, Zou L, Zhao H, Cheng C. Gene signatures
associated with genomic aberrations predict prognosis in neuroblastoma.
Cancer Communications. 2020;40: 105--118.
doi:\href{https://doi.org/10.1002/cac2.12016}{10.1002/cac2.12016}

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Prediction of \emph{PIK3CA} mutations from cancer gene expression data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jun Kang\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}},
Ahwon Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}},
Youn Soo Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}}\textsuperscript{*}\\
\bigskip
\textbf{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}}Department of Hospital Pathology, Seoul St.~Mary's Hospital, College of
Medicine, The Catholic University of Korea, Seoul, South Korea\\
\bigskip
* Corresponding author: lys9908@catholic.ac.kr\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Breast cancers with \emph{PIK3CA} mutations can be treated with
\emph{PIK3CA} inhibitors in hormone receptor-positive HER2 negative
subtypes. We applied a supervised elastic net penalized logistic
regression model to predict \emph{PIK3CA} mutations from gene expression
data. This regression approach was applied to predict modeling using the
TCGA pan-cancer dataset. Approximately 10,000 cases were available for
\emph{PIK3CA} mutation and mRNA expression data. In 10-fold
cross-validation, the model with \(\lambda\) = 0.01 and \(\alpha\) = 1.0
(Ridge regression) showed the best performance, in terms of area under
the receiver operating characteristic (AUROC). The final model was
developed with selected hyper-parameters using the entire training set.
The training set AUROC was 0.93, and the test set AUROC was 0.84. The
area under the precision-recall (AUPR) of the training set was 0.66, and
the test set AUPR was 0.39. Cancer types were the most important
predictors. Both \emph{insulin like growth factor 1 Receptor}
(\emph{IGF1R}) and the \emph{phosphatase and tensin homolog}
(\emph{PTEN}) were the most significant genes in gene expression
predictors. Our study suggests that predicting genomic alterations using
gene expression data is possible, with good outcomes.

@@

Machine learning approaches can be applied to detect genomic
alterations. Machine learning algorithms can build prediction models
from a large number of predictors, such as radiomic features {[}3{]},
pathology image {[}4{]} or gene expression data {[}5{]}. Because most
direct genomic tests are more specific and sensitive than predictive
models, machine learning approaches may have limited roles in clinical
practice, however, machine learning approaches are ideal when direct
tests are unavailable or fail.

RAS pathway activation predictions have been performed using gene
expression data {[}5{]}. Authors used data from The Cancer Genome Atlas
(TCGA), with a supervised elastic net penalized logistic regression
classifier, with stochastic gradient descent. Their model performance
was 84\% with an area under the receiver operating characteristic
(AUROC) curve, and 63\% with an area under the precision-recall (AUPR)
curve. Importantly, these authors suggested their approach could be
applied to other genomic alterations.

Breast cancer having \emph{PIK3CA} mutations can be treated using PIK3CA
inhibitors, in hormone receptor-positive HER2 negative subtypes {[}7{]}.
The \emph{PIK3CA} mutation is the second most common driver mutation
after \emph{TP53}, and is most frequently detected in endometrial
carcinoma (45\%), followed by breast invasive carcinoma (24\%), cervical
squamous cell carcinoma, endo-cervical adenocarcinoma (20\%) and colon
adenocarcinoma (16\%) {[}8{]}.

\emph{PIK3CA} encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase that
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to generate phosphatidylinositol
3,4,5-triphosphate (PIP\textsubscript{3}). The phosphatase and tensin
homolog (\emph{PTEN}) converts PIP\textsubscript{2} to
PIP\textsubscript{3} in contrast to PI3K. PIP\textsubscript{3} is a
second messenger that activates protein kinase B (AKT), which is a
serine/threonine-specific protein kinase. AKT inhibits apoptosis and
promotes cell proliferation {[}9{]}.

We applied a supervised elastic net penalized logistic regression model
to predict \emph{PIK3CA} mutations. We wanted to ascertain whether this
prediction model approach could be applied not only to RAS pathway
activation, but also to \emph{PIK3CA} mutation predictions.

@@

We used the TCGA pan-cancer dataset. TCGA archives the following; exome
sequencing, gene expression, DNA methylation, protein expression, and
clinical data from \textgreater{} 10,000 cancer samples across 33 common
cancer types. The TCGA dataset is publically available. \emph{PIK3CA}
mutation data was extracted using cgdsr rpackage {[}10{]}. Gene
expression data was downloaded from the National Cancer Institute
(NCI)'s Genomic Data Commons (GDC) website. This archives data for TCGA
(https://gdc.cancer.gov/about-data/publications/pancanatlas). Gene
expression in the TCGA pan-cancer dataset is batch-corrected with
normalization.

The target variable was \emph{PIK3CA} mutation status. \emph{PIK3CA}
status was considered positive when the case had the following
\emph{PIK3CA} variants (C420R, E542K, E545A, E545D, E545G, E545K, Q546E,
Q546R, H1047L, H1047R, H1047Y) which were the target variants of the
Therascreen \emph{PIK3CA} RGQ PCR Kit, Qiagen, Hilden, Germany. This kit
was approved as a companion diagnostics test to treat with PIK3CA
inhibitor by the United States Food and Drug Administration.

@@

To narrow down potential predictors, genes with a large median absolute
deviation (\textgreater{} third-quartiles) were selected. Thirty three
cancer type dummy variables were included in predictor variables. We
split three-quarters of the dataset into the training set and one
quarter into the test set. Yeo-Johnson transformation was performed to
correct skewness. Centering and scaling were also performed. All
preprocessing was performed using the recipe r package {[}11{]}.
Penalized logistic regression was applied to prediction modeling.
Ten-fold cross-validation with target variable stratification was
performed over the hyper-parameter grid: \(\lambda\) \{\(10^{-5}\),
\(10^{-4}\),\(10^{-3}\),\(10^{-2}\),\(10^{-1}\), \(10^{0}\)\},
\(\alpha\) \{0.0, 0.25, 0.5, 0.75, 1.0\}. Lambda (\(\lambda\)) is a
penalty scaling parameter and alpha (\(\alpha\)) is a mixing parameter
of penalty function
(\((1-\alpha)/2 \lVert\beta\rVert_2^2+\alpha\lVert \beta \rVert_1\))
{[}12{]}.

\hypertarget{assessing-model-performance}{%
\subsection{Assessing model
performance}\label{assessing-model-performance}}

Model performance was evaluated using AUROC and AUPR curve approaches.
The AUPR approach is more informative than AUROC for imbalanced datasets
{[}13{]}. The modeling process and assessing model performance were
performed with the tidymodels rpackage {[}14{]}.

@@

10,845 cases were available for both \emph{PIK3CA} mutation and mRNA
expression data. 5,128 out of 20,502 genes were included in the modeling
process, after filtering for median absolute deviation, as described in
the modeling process method. The prevalence rate for \emph{PIK3CA}
mutation was 0.11 in all cases. The \emph{PIK3CA} mutation prevalence
rate in each cancer type varied. The median prevalence rate of
\emph{PIK3CA} mutation for each cancer type was 0.03 (range 0--0.33)
(Figure 1).

@@

Because \emph{PIK3CA} mutation prevalence varied across cancer types,
the performance of each cancer type was investigated. The AUROC and AUPR
were positively correlated between the training sets and test sets in
cancer type sub-analysis (Figure 2B). The AUPR was high in cancer types
with high \emph{PIK3CA} mutation rates such as colon, breast and uterus
cancer types. The AUROC did not correlate with \emph{PIK3CA} mutation
rates of each cancer type (Figure 2C).

@@

The top 30 important predictors are shown (Figure 3). The coefficient is
the parameter of the predictor which represents the effect of the
predictor on prediction. \emph{Insulin like growth factor 1 Receptor}
(\emph{IGF1R}) mRNA expression was the strongest negative predictor, and
\emph{PTEN} was the strongest positive predictor. Both \emph{IGF1R} and
\emph{PTEN} are key players in the tyrosine kinase pathway {[}9,15{]}.
The cancer types were important predictors. Some cancer types including
uterine carcinosarcoma (UCS), bladder urothelial carcinoma (BLCA),
pancreatic adenocarcinoma (PAAD), lymphoid neoplasm diffuse large B-cell
lymphoma (DLBC) were the strongest predictors.

@@

Our model showed good performance in predicting \emph{PIK3CA} mutations
in various cancer types. Our data suggested that the supervised elastic
net penalized logistic regression model could be applied not only to the
RAS activation pathway, but also to other genomic alterations. Both the
RAS activation pathway and \emph{PIK3CA} mutations are key, common
cancer genomic alterations. Because they exert significant effect on
gene expression in cancer cells, prediction from gene expression data
can be good. However, the supervised elastic net penalized logistic
regression model cannot be generalized or applied to other genomic
alterations which have have a weak effect on gene expression.

Prediction modeling from the TCGA pan-cancer dataset can be limiting in
terms of data preprocessing. The gene expression data is processed by
between-sample normalization to remove batch effects. If the model has
been trained from between-sample normalization, a new sample cannot be
exactly processed with normalization which was done on trainset. A model
based on gene expression from the TCGA pan-cancer dataset has limitation
in terms of data preprocessing. It is necessary to develop a processing
method that is independent of a dataset, to apply gene expression data
to the prediction model.

Our \emph{PIK3CA} prediction model was similar to the RAS activation
prediction model in terms of AUROC (0.84). However the AUPR of our model
was lower than the RAS activation model (0.39 versus 0.63). The reason
for our lower AUPR may be explained by an imbalanced dataset that has
the low prevalence rate of \emph{PIK3CA} mutations {[}5{]}. The model
for RAS activation trained with cancer types with more than 0.05
prevalence of RAS activation to avoid imbalance classification problem.
We included all cancer types in our modeling process. The lower
prevalence rate of target variables meant our dataset had a lower AUPR
baseline. In the sub-analysis performance of each cancer type, the
cancer types with higher \emph{PIK3CA} mutation rates showed better
AUPRs.

Our model included cancer types as predictors, and they were stronger
predictors than gene expression. The varying prevalence of PIC3CA
mutations across cancer types may be a reason for the strong predictive
power of cancer types.

Some significant gene expression predictors were closely related to the
PTEN-PI3K pathway. \emph{PTEN} and \emph{IGFR1R} were the strongest gene
expression predictors, which has negative and positive predictive
powers. \emph{IGF1R} is a tyrosine kinase receptor that activates PI3K
{[}15{]}, and \emph{PTEN} is an important regulator of
PIP\textsubscript{3} by dephosphorylating PIP\textsubscript{3} {[}9{]}.

Several studies have attempted to predict genomic alterations from gene
expression data {[}6,16{]}. A study investigated \emph{PIK3CA} mutation
predictions using gene-expression signatures which is a sum of the
average of the logarithmic gene expression. The model showed good
performance AUROC 0.71 in an independent test set {[}6,16{]}. Another
study predicted copy number alterations with gene expression, using a
multinomial logistic regression model with least absolute shrinkage and
selection operator (LASSO) parameters {[}17{]}. The prediction of the
1p/19q codeletion was very good, with an AUROC of 0.997, and gene-level
predictions were good, with an AUROC of 0.75 {[}17{]}. A logistic
regression model was used for \emph{MYCN Proto-Oncogene, BHLH
Transcription Factor} (\emph{MYCN}) gene amplification in neuroblastoma
{[}18{]}.

Our study suggested that the prediction of genomic alterations using
gene expression data was possible, with good performance. However,
improved performances are required for clinical tests, and the
standardization of generation processing of gene expression data is also
needed.

@@

\begin{itemize}
\item
  Figure 1. Prevalence rate of \emph{PIK3CA} mutations across cancer
  types. Cancer type abbreviations are explained in the S1 Appendix.
\item
  Figure 2. Summary of modeling results. (A) Left: receiver operating
  characteristic (ROC) curve. Right: precision-recall (PR) curve of
  training set and test set. The horizontal green line is the
  \emph{PIK3CA} mutation rate (0.11) (B) Correlation between training
  set and test set of the area under the receiver operating
  characteristic curve (AUROC), and the area under the precision-recall
  curve (AUPR) among cancer types. The gray band is the 95\% confidence
  interval. Abbreviations are explained in the S1 Appendix. (C)
  Correlations between the \emph{PIK3CA} mutation rate of the AUROC, and
  the AUPR.
\item
  Figure 3. Coefficient model. (A) Top 30 high mRNA coefficients. (B)
  Cancer type coefficients. Cancer types abbreviations are explained in
  the S1 Appendix.
\end{itemize}

@@

\leavevmode\hypertarget{ref-andre_2019_alpelisib_nengljmeda}{}%
7. André F, Ciruelos E, Rubovszky G, Campone M, Loibl S, Rugo HS, et al.
Alpelisib for PIK3CA-Mutated, Hormone ReceptorPositive Advanced Breast
Cancer. New England Journal of Medicine. Massachusetts Medical Society;
2019;380: 1929--1940.
doi:\href{https://doi.org/10.1056/NEJMoa1813904}{10.1056/NEJMoa1813904}

\leavevmode\hypertarget{ref-baileyComprehensiveCharacterizationCancer2018}{}%
8. Bailey MH, Tokheim C, Porta-Pardo E, Sengupta S, Bertrand.
Comprehensive Characterization of Cancer Driver Genes and Mutations.
Cell. 2018;173: 371--385.e18.
doi:\href{https://doi.org/10.1016/j.cell.2018.02.060}{10.1016/j.cell.2018.02.060}

\leavevmode\hypertarget{ref-cantleyPhosphoinositide3KinasePathway2002}{}%
9. Cantley LC. The Phosphoinositide 3-Kinase Pathway. Science. American
Association for the Advancement of Science; 2002;296: 1655--1657.
doi:\href{https://doi.org/10.1126/science.296.5573.1655}{10.1126/science.296.5573.1655}

\leavevmode\hypertarget{ref-jacobsenCgdsrRbasedAPI2019}{}%
10. Jacobsen A, Luna A. Cgdsr: R-based API for accessing the MSKCC
cancer genomics data server (CGDS). 2019.

\leavevmode\hypertarget{ref-kuhnRecipesPreprocessingTools2020}{}%
11. Kuhn M, Wickham H. Recipes: Preprocessing tools to create design
matrices. 2020.

\leavevmode\hypertarget{ref-friedmanRegularizationPathsGeneralized2010}{}%
12. Friedman J, Hastie T, Tibshirani R. Regularization Paths for
Generalized Linear Models via Coordinate Descent. J Stat Softw. 2010;33:
1--22.

\leavevmode\hypertarget{ref-saitoPrecisionRecallPlotMore2015}{}%
13. Saito T, Rehmsmeier M. The Precision-Recall Plot Is More Informative
than the ROC Plot When Evaluating Binary Classifiers on Imbalanced
Datasets. PLOS ONE. Public Library of Science; 10: e0118432.
doi:\href{https://doi.org/10.1371/journal.pone.0118432}{10.1371/journal.pone.0118432}

\leavevmode\hypertarget{ref-kuhnTidymodelsEasilyInstall2020}{}%
14. Kuhn M, Wickham H. Tidymodels: Easily install and load the
'tidymodels' packages. 2020.

\leavevmode\hypertarget{ref-leroithInsulinlikeGrowthFactor2003}{}%
15. LeRoith D, Roberts CT. The insulin-like growth factor system and
cancer. Cancer Letters. 2003;195: 127--137.
doi:\href{https://doi.org/10.1016/S0304-3835(03)00159-9}{10.1016/S0304-3835(03)00159-9}

\leavevmode\hypertarget{ref-cizkovaGeneExpressionProfiling2010}{}%
16. Cizkova M, Cizeron-Clairac G, Vacher S, Susini A, Andrieu C,
Lidereau R, et al. Gene Expression Profiling Reveals New Aspects of
PIK3CA Mutation in ERalpha-Positive Breast Cancer: Major Implication of
the Wnt Signaling Pathway. PLOS ONE. Public Library of Science; 5:
e15647.
doi:\href{https://doi.org/10.1371/journal.pone.0015647}{10.1371/journal.pone.0015647}

\leavevmode\hypertarget{ref-muCNAPEMachineLearning2019}{}%
17. Mu Q, Wang J. CNAPE: A Machine Learning Method for Copy Number
Alteration Prediction from Gene Expression. IEEE/ACM Transactions on
Computational Biology and Bioinformatics. 2019; 1--1.
doi:\href{https://doi.org/10.1109/TCBB.2019.2944827}{10.1109/TCBB.2019.2944827}

\leavevmode\hypertarget{ref-heGeneSignaturesAssociated2020}{}%
18. He X, Qin C, Zhao Y, Zou L, Zhao H, Cheng C. Gene signatures
associated with genomic aberrations predict prognosis in neuroblastoma.
Cancer Communications. 2020;40: 105--118.
doi:\href{https://doi.org/10.1002/cac2.12016}{10.1002/cac2.12016}
